The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-10-07 paper.tex
Revisao ate 4.4

\begin{figure}
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  emit e => 1;
with
  v = await e;
  _printf("%d\n",v);
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  e_ = 1;
  emit e;
with
  await e;
  v = e_;
  _printf("%d\n",v);
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
<...> ; mem ; emit(e)
        or
await(e) ; mem ; mem
\end{lstlisting}
\end{minipage}
\caption{
Two-step translation from concrete to abstract \code{emit} and \code{await} 
expressions.
The concrete code in the left communicates the value \code{1} from the 
\code{emit} to the \code{await}.
The abstract code uses a shared variable to hold the value.
\label{lst.map.emit.await}
}
\end{figure}

@@

To encompass first-class timers, we introduce a special external \code{DT} 
event that is intercalated with each other event occurrence in an application 
(e.g.  \emph{e1, e2}):

@@

The \code{DT} event has an associated variable \code{DT\_} carrying the 
wall-clock time elapsed between the two occurrences, as depicted by the 
two-step translation of Figure~\ref{lst.map.timers.2}.
In the concrete program in the left, the variable \code{dt} will hold the 
residual delta time (as described in Section~\ref{sec.ceu.timers}) after 
awaking from the timer.
In the first step of the translation, we expand the \code{await 10ms} to a 
\code{loop} that decrements the elapsed number of microseconds for each 
occurrence of \code{DT}.
When the variable \code{tot} reaches zero, we escape the \code{loop} setting 
the variable \code{dt} to contain the appropriate delta.
In the last step, we convert the program to the abstract syntax.

\begin{figure}[b]
\begin{minipage}[t]{0.30\linewidth}
\begin{lstlisting}
dt = await 10ms;
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.37\linewidth}
\begin{lstlisting}
var int tot = 10000; // 10ms
loop do
    await DT;
    tot = tot - DT_;
    if tot <= 0 then
        dt = tot;
        break;
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.30\linewidth}
\begin{lstlisting}
mem;
loop(
    await(DT);
    mem;
    if mem then
        mem;
        break
    else
        nop
)
\end{lstlisting}
\end{minipage}
\caption{
Two-step translation from concrete to abstract timer.
\label{lst.map.timers.2}
}
\end{figure}

@@

The biggest mismatch between concrete and abstract \CEU is regarding 
finalization blocks, which require more complex modifications in the program
for a proper mapping using the $fin$ expressions.
In the three-step translation of Figure~\ref{lst.map.fin}, we start with a 
concrete program (\code{CODE-1}) that uses a \code{finalize} to safely 
\code{\_release} the reference to \code{ptr} kept after the call to 
\code{\_hold}.
In the translation, we first need to catch the \code{do-end} termination to run 
the finalization code.
For this, we translate the block into a \code{par/or} (\code{CODE-2}) with the 
original body in parallel with a $fin$ to run the finalization code.
This way, the $fin$ body executes whenever the \code{par/or} terminates, either 
normally (after the \code{await B}) or aborted from an outer composition.
%
However, the $fin$ still (incorrectly) executes even if the call to 
\code{\_hold} is not reached in the body due to an abort before awaking from 
the \code{await A}.
%
To deal with this issue, for each $fin$ we need a corresponding flag to keep 
track of code that needs to be finalized (\code{CODE-3}).
%
The flag is initially set to false, avoiding the finalization code to execute.
Only after the call to \code{\_hold} that we set the flag to true and enable 
the $fin$ body to execute.
%
The complete translation substitutes the side-effect operations with $mem$ 
expressions (\code{CODE-4}).

@@

This phase detects inconsistencies in \CEU programs, such as unbounded loops, 
suspicious accesses to shared memory, and also ``classical'' semantic analysis, 
such as building a symbol table for checking variable declarations.
%
This phase outputs code in \C, given the way \CEU is tied to \C by design.
%
Some type checking is delayed to the last phase to take advantage of 
\code{gcc}'s error handling.
Therefore, we annotate the \C output file with \code{\#line} pragmas matching 
the original file in \CEU.

\item[Code Generation]

@@

\CEU favors a fine-grained use of trails, being common the use of trails that 
await a single event.
For this reason, \CEU{} does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to \C is straightforward~\cite{wsn.osm,wsn.ocram}: memory 
for blocks in sequence are packed in a \code{struct}, while blocks in parallel, 
in a \code{union}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at a given time.
A given position in the memory may hold different data (with variable sizes) 
during runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
memory layout.
%
Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
with the same name can be distinguished.
%
The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
\code{MEM.b\_2} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags related to the 
parallel composition, which also reside in reusable slots in the static memory.

@@

At any given moment, a trail can be awaiting in one of the following states: 
\code{INACTIVE}, \code{STACKED}, \code{FIN}, or in any of the events defined in 
the program:

\begin{lstlisting}
enum {
    INACTIVE = 0,
    STACKED,
    FIN,
    EVT_A,      // input void A;
    EVT_e,      // event int e;
    <...>       // other events
}
\end{lstlisting}

All terminated or not-yet-started trails stay in the \code{INACTIVE} state and 
are ignored by the scheduler.
%
A \code{STACKED} trail holds an associated stack level and is delayed until the 
scheduler runtime reaches that level again.
%
A \code{FIN} trail represents a hanged finalization block which is only 
scheduled when its corresponding block goes out of scope.
%
A trail waiting for an event stays in the state of the corresponding event, 
also holding the minimum sequence number (\emph{seqno}) in which it can awake.
%
In concrete terms, a trail is represented by the following \code{struct}:

@@

The field \code{evt} holds the state of the trail (or the event it is 
awaiting); the field \code{lbl} holds the entry point in the code to execute 
when the trail is scheduled; the third field depends on the \code{evt} field 
and may hold the \code{seqno} for an event, or the stack level \code{stk} for a
\code{STACKED} state.

@@

In the final generated code in \C, each trail label representing an entry point 
becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in the left of the figure, the compiler extracts the entry 
points and associated trails, e.g., the label \code{Awake\_e} will execute on 
\code{TRAIL-0} (ln. 7).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
in the middle of the figure).
The state of trails translate to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (ln. 12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(ln. 13), while all others are set to \code{INACTIVE} (in the example, only 
one, in ln. 14).

@@

Regarding abortion and finalization, when a \code{par/or} terminates, the 
scheduler makes a \emph{broadcast} pass for the \code{FIN} event, but limited 
to the range of trails covered by the terminating \code{par/or}.
Trails that do not match the \code{FIN} are set to \code{INACTIVE}, as they 
have to be aborted.
Given that trails in parallel are allocated in subsequent slots in the static 
vector \code{TRLS}, this pass only aborts the desirable trails.
The subsequent \emph{dispatch} pass executes the finalization code.
%Given that finalization blocks cannot contain \code{await} statements, the 
%whole process is guaranteed to terminate in bounded time.
Escaping a \code{loop} that contains parallel compositions also triggers the 
same abortion process.

>>>>

\begin{figure}[t]
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  emit e => 1;
with
  v = await e;
  _printf("%d\n",v);
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  e_ = 1;
  emit e;
with
  await e;
  v = e_;
  _printf("%d\n",v);
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
<...> ; mem ; emit(e)
        or
await(e) ; mem ; mem
\end{lstlisting}
\end{minipage}
\caption{
Two-step translation from concrete to abstract \code{emit} and \code{await} 
expressions.
The concrete code in the left communicates the value \code{1} from the 
\code{emit} to the \code{await}.
The abstract code uses a shared variable to hold the value.
\label{lst.map.emit.await}
}
\end{figure}

@@

To encompass first-class timers, we introduce a special external event 
\code{DT} that is intercalated with each other event occurrence in an 
application (e.g.  \emph{e1, e2}):

@@

The event \code{DT} has an associated variable \code{DT\_} carrying the 
wall-clock time elapsed between two occurrences in sequence, as depicted by the 
two-step translation of Figure~\ref{lst.map.timers.2}.
In the concrete program in the left, the variable \code{dt} holds the residual 
delta time (as described in Section~\ref{sec.ceu.timers}) after awaking from 
the timer.
In the first step of the translation, we expand the \code{await 10ms} to a 
\code{loop} that decrements the elapsed number of microseconds for each 
occurrence of \code{DT}.
When the variable \code{tot} reaches zero, we escape the \code{loop} setting 
the variable \code{dt} to contain the appropriate delta.
In the last step, we convert the program to the abstract syntax.

\begin{figure}[b]
\begin{minipage}[t]{0.30\linewidth}
\begin{lstlisting}
dt = await 10ms;
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.37\linewidth}
\begin{lstlisting}
var int tot = 10000; // 10ms
loop do
    await DT;
    tot = tot - DT_;
    if tot <= 0 then
        dt = -tot;
        break;
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.30\linewidth}
\begin{lstlisting}
mem;
loop(
    await(DT);
    mem;
    if mem then
        mem;
        break
    else
        nop
)
\end{lstlisting}
\end{minipage}
\caption{
Two-step translation from concrete to abstract timer.
\label{lst.map.timers.2}
}
\end{figure}

@@

The biggest mismatch between concrete and abstract \CEU is regarding the 
\code{finalize} blocks, which require more complex modifications in the program
for a proper mapping using $fin$ expressions.
In the three-step translation of Figure~\ref{lst.map.fin}, we start with a 
concrete program (\code{CODE-1}) that uses a \code{finalize} to safely 
\code{\_release} the reference to \code{ptr} kept after the call to 
\code{\_hold}.
In the translation, we first need to catch the outermost \code{do-end} 
termination to run the finalization code.
For this, we translate the block into a \code{par/or} (\code{CODE-2}) with the 
original body in parallel with a $fin$ expression to run the finalization code.
Note that the $fin$ has no transition rules in the semantics, keeping the 
\code{par/or} alive.
This way, the $fin$ body only executes when the \code{par/or} terminates either 
normally (after the \code{await B}), or aborted from an outer composition.
%
However, the $fin$ still (incorrectly) executes even if the call to 
\code{\_hold} is not reached in the body due to an abort before awaking from 
the \code{await A}.
%
To deal with this issue, for each $fin$ we need a corresponding flag to keep 
track of code that needs to be finalized (\code{CODE-3}).
%
The flag is initially set to false, avoiding the finalization code to execute.
Only after the call to \code{\_hold} that we set the flag to true and enable 
the $fin$ body to execute.
%
The complete translation substitutes the side-effect operations with $mem$ 
expressions (\code{CODE-4}).

@@

This phase detects inconsistencies in \CEU programs, such as unbounded loops, 
suspicious accesses to shared memory, and also performs ``classical'' semantic 
analysis, such as building a symbol table for checking variable declarations.
%
For the (standard) \C back end, this phase outputs code in \C.
%
Some type checking is delayed to the last phase to take advantage of 
\code{gcc}'s error handling.
Therefore, we annotate the \C output file with \code{\#line} pragmas matching 
the original file in \CEU.

\item[Final Generation]

@@

\CEU favors a fine-grained use of trails, being common to use trails that await 
a single event.
For this reason, \CEU{} does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to \C is straightforward~\cite{wsn.osm,wsn.ocram}: memory 
for blocks in sequence are packed in a \code{struct}, while blocks in parallel, 
in a \code{union}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at a given time.
A given position in the memory may hold different data (with variable sizes) 
during runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
memory layout.
%
Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
with the same name can be distinguished.
%
The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
\code{MEM.b\_2} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags related to the 
parallel composition, which also reside in reusable slots in the static memory.

@@

At any given moment, a trail can be awaiting in one of the following states: 
\code{INACTIVE}, \code{STACKED}, \code{FINALIZE}, or in any of the events 
defined in the program:

\begin{lstlisting}
enum {
    INACTIVE = 0,
    STACKED,
    FINALIZE,
    EVT_A,      // input void A;
    EVT_e,      // event int e;
    <...>       // other events
}
\end{lstlisting}

All terminated or not-yet-started trails stay in the \code{INACTIVE} state and 
are ignored by the scheduler.
%
A \code{STACKED} trail holds an associated stack level and is delayed until the 
scheduler runtime reaches that level again.
%
A \code{FINALIZE} trail represents a hanged finalization block which is only 
scheduled when its corresponding block goes out of scope.
%
A trail waiting for an event stays in the state of the corresponding event, 
also holding the minimum sequence number (\emph{seqno}) in which it can awake.
%
In concrete terms, a trail is represented by the following \code{struct}:

@@

The field \code{evt} holds the state of the trail (or the event it is 
awaiting); the field \code{lbl} holds the entry point in the code to execute 
when the trail segment is scheduled; the third field depends on the \code{evt} 
field and may hold the \code{seqno} for an event, or the stack level \code{stk} 
for a
\code{STACKED} state.

@@

In the final generated code in \C, each trail segment label representing an 
entry point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in the left of the figure, the compiler extracts the entry 
points and associated trails, e.g., the label \code{Awake\_e} will execute on 
\code{TRAIL-0} (ln. 7).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
in the middle of the figure).
The state of trails translate to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (ln. 12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(ln. 13), while all others are set to \code{INACTIVE} (in the example, only 
one, in ln. 14).

@@

Regarding abortion and finalization, when a \code{par/or} terminates, the 
scheduler makes a \emph{broadcast} pass for the \code{FINALIZE} event, but 
limited to the range of trails covered by the terminating \code{par/or}.
Trails that do not match the \code{FINALIZE} are set to \code{INACTIVE}, as 
they have to be aborted.
Given that trails in parallel are allocated in subsequent slots in the static 
vector \code{TRLS}, this pass only aborts the desirable trails.
The subsequent \emph{dispatch} pass executes the finalization code.
%Given that finalization blocks cannot contain \code{await} statements, the 
%whole process is guaranteed to terminate in bounded time.
Escaping a \code{loop} that contains parallel compositions also triggers the 
same abortion process.
